---
title: "QB_final_project"
author: "Lana Bolin"
date: "2/6/2019"
output: html_document
---

---
title: "Bolin, Caple, and Crawley independent project (QB2019)"
author: "Lana Bolin"
date: "1/31/2019"
output:
  pdf_document: default
  html_document: default
---

# Setup

```{r, warning=F, message=F}
rm(list = ls())

package.list <- c('vegan', 'dplyr', 'magrittr', 'ggplot2', 'BiodiversityR', 'indicspecies')
for(package in package.list){
  if(!require(package, character.only = TRUE, quietly = TRUE)){
    install.packages(package)
    library(package, character.only = TRUE)
  }
}

```


Nativity codes are:   
0 = native to Indiana  
1 = native to the US  
2 = exotic  
3 = invasive  

Check how many of each nativity code we have using dplyr
```{r}
setwd("~/GitHub/QB19_IndependentProject/")
crawley.full <- read.csv("Crawley.csv")

crawley.full %>%
  group_by(Nativity.Code) %>%
  summarize(grpcount = n())

# Those 8 empty ones are just empty rows at the bottom of the df. Let's get rid of those, as well as the two empty columns at the end
crawley.full <- crawley.full[-c(293:300), -c(24, 25)]
crawley.df <- crawley.full[, -c(1:3)]  # Get rid of non-incidence columns
crawley.species <- crawley.full$Species  # Save species as a vector, in case we need it
crawley <- as.matrix(t(crawley.df))  

crawley.sbys <- matrix(crawley, ncol = ncol(crawley), dimnames = NULL)

```


# Plots

## Richness vs. Invasive present / Invasive absent in community

Make site-by-species matrix where nativity code is "0" for Native, "1" for Invasive/Exotic
```{r}
# We want to separate into two df's: one for sites that have at least one invasive, and one for sites that don't.

# To do that, we make a vector of invasive species

crawley.inv <- crawley.full[crawley.full$ Nativity.Code == "3", ]

```

#### Every site has at least one invasive, so I can't make my plot!



## Mackenzie's initial idea: plot percent invasive species vs. species richness

```{r}
# subset data into separate data frames for each nativity code
nativeIN <- subset(crawley.full, Nativity.Code == "0")
nativeUS <- subset(crawley.full, Nativity.Code == "1")
exotic <- subset(crawley.full, Nativity.Code == "2")
invasive <- subset(crawley.full, Nativity.Code == "3")
unknown <- subset(crawley.full, Nativity.Code == "UNK")

# sum columns to find observed species richness for each site; repeat for each nativity code
nativeIN.sr <- colSums(nativeIN[ , 4:23])
nativeUS.sr <- colSums(nativeUS[ , 4:23])
exotic.sr <- colSums(exotic[ , 4:23])
invasive.sr <- colSums(invasive[ , 4:23])
unknown.sr <- colSums(unknown[ , 4:23])

# calculate total native richness
native.sr <- nativeIN.sr + nativeUS.sr

# calculate total observed species richness
total.sr <- colSums(crawley.full[ , 4:23])

# this was a sanity check to make sure everything has worked so far
# total.osr <- nativeIN.sr + nativeUS.sr + exotic.sr + invasive.sr + unknown.sr
# total.sr == total.osr

# plot richness of invasive species vs total species richness
plot(invasive.sr, total.sr, ylim = range(0:70), xlim = range(0:15), 
     xlab = "Number of Invasive Species Present", ylab = "Total Observed Species Richness")
# add fit lines
abline(lm(total.sr~invasive.sr), col="blue") # regression line (y~x) 

# plot richness of invasive species vs native species richness
plot(invasive.sr, native.sr, ylim = range(0:70), xlim = range(0:15), 
     xlab = "Number of Invasive Species Present", ylab = "Number of Native Species Present", main = "Observed Richness of Native and Invasive Plants")
# add fit lines
abline(lm(native.sr~invasive.sr), col="blue") # regression line (y~x) 


# plot richness of native species richness vs invasive species 
plot(native.sr, invasive.sr, ylim = range(0:20), xlim = range(0:80), 
     ylab = "Number of Invasive Species Present", xlab = "Number of Native Species Present", main = "Observed Richness of Native and Invasive Plants")
# add fit lines
abline(lm(invasive.sr~native.sr), col="blue") # regression line (y~x) 

summary(lm(native.sr~invasive.sr))

```

> Ok, so there's a pretty clear positive trend for both graphs, which is pretty neat-- so from this data, it looks like adding invasives simply adds species, and there isn't evidence that the invasives are decreasing native or total species richness (the opposite, in fact, which is logical for the first graph but really interesting for the second graph)


## Beta diversity of sites

```{r}
# loading transposed site by species matrix with nativity/planting codes removed
sbs.factors <- read.csv("sbs_factors.csv", header = TRUE)
sbs <- sbs.factors[ , -c(1:4)]

## this was me trying to transpose the matrix, set the first row as column names and then giving up because the data was no longer considered numeric?
## i'm really just keeping this in the hopes that one day i'll be able to figure it out, we don't need it, feel free to delete
#sbs <- crawley.full[-c(293:300) ,-c(2, 3, 24, 25)]
#sbs <- t(sbs)
#colnames(sbs) <- sbs[1, ]
#sbs <- sbs[-1, -c(293:300)]
 

# sorted data by site factors
mulch.sbs <- sbs.factors[order(sbs.factors$MULCH),]
litter.sbs <- sbs.factors[order(sbs.factors$LITTER),]
weediness.sbs <- sbs.factors[order(sbs.factors$WEEDINESS),]

# calculated Sorenson similarity
msbs.ds <- vegdist(mulch.sbs[ , -c(1:4)], method = "bray", binary = TRUE)
lsbs.ds <- vegdist(litter.sbs[ , -c(1:4)], method = "bray", binary = TRUE)
wsbs.ds <- vegdist(weediness.sbs[ , -c(1:4)], method = "bray", binary = TRUE)


# heat maps of site similarity
require(viridis)
levelplot(as.matrix(msbs.ds), aspect = "iso", col.regions = magma,
                    xlab = "GreenSpace Site (Sort: Mulch Abundance)", ylab = "GreenSpace Site (Sort: Mulch Abundance)", scales = list(cex = 0.5),
                    main = "Sorenson Distance")
levelplot(as.matrix(lsbs.ds), aspect = "iso", col.regions = magma,
                    xlab = "GreenSpace Site (Sort: Litter Abundance)", ylab = "GreenSpace Site (Sort: Litter Abundance)", scales = list(cex = 0.5),
                    main = "Sorenson Distance")
levelplot(as.matrix(wsbs.ds), aspect = "iso", col.regions = magma,
                    xlab = "GreenSpace Site (Sort: Weediness Index)", ylab = "GreenSpace Site (Sort: Weediness Index)", scales = list(cex = 0.5),
                    main = "Sorenson Distance")

```

> Not expecting to find any patterns with site similarity honestly (and it doesn't look like there are any!), but it's cool to see that they're overall pretty dissimilar. Wonder if that changes looking only at planted species?
>Not sure if the labels help, but the heat maps just differ based on the sorting of the dataset; the values are ascending, so right to left and down to up is least mulch/litter/weeds to most.

## Planted species vs. volunteer species

```{r}
# ignore this for now, I'm just playin' 
planted <- subset(crawley.full, Planted.or.Volunteer == "1")
volunteer <- subset(crawley.full, Planted.or.Volunteer == "2")
planted.SR <- colSums(planted[ , 4:23])
volunteer.SR <- colSums(volunteer[ , 4:23])

plot(planted.SR, volunteer.SR, ylim = c(0, 80), xlim = c(0, 30),
     xlab = "Planted Species Richness",
     ylab = "Volunteer Species Richness", las = 1)

```


## PERMANOVA using site types and locations 

```{r}
sbs.ds <- vegdist(sbs, method = "bray", binary = TRUE) 
sbs.pcoa <- cmdscale(sbs.ds, eig = TRUE, k = 3)
explainvar1.sbs <- round(sbs.pcoa$eig[1] / sum (sbs.pcoa$eig), 3) * 100
explainvar2.sbs <- round(sbs.pcoa$eig[2] / sum (sbs.pcoa$eig), 3) * 100
explainvar3.sbs <- round(sbs.pcoa$eig[3] / sum (sbs.pcoa$eig), 3) * 100
sum.eigsbs <- sum(explainvar1.sbs, explainvar2.sbs, explainvar3.sbs)
sum.eigsbs

par(mar = c(4, 4, 1, 1) + 0.1) 
plot(sbs.pcoa$points[, 1], sbs.pcoa$points[, 2],
     ylim = c(-1, 1), xlim = c(-1, 1),
     xlab = paste("PCoA 1 (", explainvar1.sbs, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2.sbs, "%)", sep = ""),
     pch = 16, cex = 0.8, type = "n", cex.lab = 0.8, cex.axis = 0.8, axes = FALSE)
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 0.8, las = 1) 
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 0.8, las = 1) 
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
points(sbs.pcoa$points[, 1], sbs.pcoa$points[, 2], 
    pch = 19, cex = 2, bg = "gray", col = "gray")
text(sbs.pcoa$points[, 1], sbs.pcoa$points[, 2], cex = 0.8,
    labels = rownames(sbs))

# add species labels to PCoA
sbs.pcoa <- add.spec.scores(sbs.pcoa, sbs, method = "pcoa.scores")
text(sbs.pcoa$cproj[ , 1], sbs.pcoa$cproj[ , 2], cex = 0.5,
     labels = row.names(sbs.pcoa$cproj), col = "blue")

# site type (i.e. school, residential, etc.)
site.type <- c(rep("Business", 1), rep ("Residential", 1), rep("School", 1), rep("Residential", 1), rep("Business", 1), rep("Other", 1), rep ("Business", 1), rep("Other", 2), rep("Residential", 1), rep("School", 2), rep ("Business", 1), rep("School", 1), rep("Residential", 2), rep("School", 1), rep("Other", 1), rep("Business", 1), rep("Residential", 1))
adonis(sbs ~ site.type, method = "bray", permutations = 999)

# median income classification
region.SES <- c(rep("Mid", 1), rep("Low", 1), rep("Mid", 1), rep("Low", 2), rep("Mid", 1), rep("Low", 1), rep("Upper", 1),  rep("Low", 2), rep("Mid", 2), rep("Upper", 1), rep("Low", 3), rep("Upper", 1), rep("Mid", 1), rep("Low", 1), rep("Mid", 1)) 
adonis(sbs ~ region.SES, method = "bray", permutations = 999)

# weediness
weediness <- c(rep("Absent", 2), rep("Low", 6), rep("Medium", 10), rep("High", 2))
adonis(sbs ~ weediness, method = "bray", permutations = 999)

```

```{r}
## Percent Invasive vs. Time Since Planting
# Setup
setwd("~/GitHub/QB19_IndependentProject/")
crawley.full <- read.csv("Crawley.csv")
crawley.site.factors <- read.csv("site_factors.csv")
time.since.planting <- crawley.site.factors$Years.between.planting.and.census  # this is an integer, for now

crawley.full %>%
  group_by(Nativity.Code) %>%
  summarize(grpcount = n())

crawley.full <- crawley.full[-c(293:300), -c(24, 25)]

# Calculate percent invasive for each site
percent.invasive <- colSums(crawley.df) / nrow(crawley.df)


plot(x = time.since.planting, y = percent.invasive,
     xlab = "Years Since Planting",
     ylab = "Percent Invasive Species in Site")
abline(lm(percent.invasive ~ time.since.planting), col="blue")

summary(lm(percent.invasive ~ time.since.planting))  # P = 0.23
```

```{r}
# mulch
mulch <- c(rep("None", 3), rep("Low", 6), rep("Medium", 7), rep("High", 4))
adonis( sbs ~ mulch, method = "bray", permutations = 999)

# didn't do litter- realized that only two sites had any litter at all. Explains why the heat map looks so homogenous! 
# also neither of the PERMANOVAs were significant

```

```{r}
#dbrda plot with site factors
site.factors <- sbs.factors[ ,2:4]
sbs.dbrda <- dbrda(sbs.ds ~ ., as.data.frame(site.factors))
ordiplot(sbs.dbrda)

permutest(sbs.dbrda, permutations = 999)
envfit(sbs.dbrda, site.factors, perm = 999)

#mantel test? not sure if done right.
sitefact.ds <- vegdist(site.factors, method = "euclidean")
mantel(sbs.ds,sitefact.ds)
```


Indicator species
```{r}
# test hypothesis: there are relatively few species driving community differences

# transpose data to create a site-by-species matrix
sbs <- t(crawley.full[ , 4:23])
sbs <- as.data.frame(sbs)
colnames(sbs) <- crawley.full[ , 1]

# create "factors" vector
richness <- rowSums(sbs)

levels <- function(x){
  if(x <= 45){
    return("L")
  } else if(x > 45 & x <= 70){
    return("M")
  } else{
    return("H")
  }
}

richness.level <- sapply(richness, levels)


# run PERMANOVA with adonis function
adonis(sbs ~ richness.level, method = "bray", permutations = 999)

# identify indicator species
indval <- multipatt(sbs, cluster = richness.level, func = "IndVal.g", control = how(nperm = 999))
summary(indval)

# calculate phi coefficient of association to examine habitat preference
sbs.rel <- decostand(sbs, method = "total")
phi <- multipatt(sbs, cluster = richness.level, func = "r.g", control = how(nperm = 999))
summary(phi)
```



Ideas for further analysis:
-incorporate plot size
-indicator species
-planted species vs volunteer species
