---
title: "QB_final_project"
author: "Lana Bolin"
date: "2/6/2019"
output: html_document
---

---
title: "Bolin, Caple, and Crawley independent project (QB2019)"
author: "Lana Bolin"
date: "1/31/2019"
output:
  pdf_document: default
  html_document: default
---

# Setup

```{r, warning=F, message=F}
rm(list = ls())
getwd()
setwd("~/GitHub/QB19_IndependentProject/")
require("vegan")
require("dplyr")
require("magrittr")
require("ggplot2")

crawley.full <- read.csv("Crawley.csv")
```


Nativity codes are:   
0 = native to Indiana  
1 = native to the US  
2 = exotic  
3 = invasive  

Check how many of each nativity code we have using dplyr
```{r}
crawley.full %>%
  group_by(Nativity.Code) %>%
  summarize(grpcount = n())

# Those 8 empty ones are just empty rows at the bottom of the df. Let's get rid of those, as well as the two empty columns at the end
crawley.full <- crawley.full[-c(293:300), -c(24, 25)]

```


# Plots

## Richness vs. Invasive present / Invasive absent in community

Make site-by-species matrix where nativity code is "0" for Native, "1" for Invasive/Exotic
```{r}
# We want to separate into two df's: one for sites that have at least one invasive, and one for sites that don't.

# To do that, we make a vector of invasive species

crawley.inv <- crawley.full[crawley.full$ Nativity.Code == "3", ]

```

#### Every site has at least one invasive, so I can't make my plot!






